\textbf{SPECFEM3D Cartesian}\\ \textbf{User Manual}

### 2015-10-20 doc/USER_MANUAL/05_running_the_solver.tex
added Quentin Brissaud's note on the acoustic formulation to the users manual

>>>>

%------------------------------------------------------------------------------------------------%

%------------------------------------------------------------------------------------------------%

\chapter{Note on the SPECFEM acoustic formulation}

{\Large Note from Quentin Brissaud, Dimitri Komatitsch and Rapha\"el Garcia, October 2015}

\section{About the potential formulation used in SPECFEM for pure acoustic waves (i.e. in the absence of gravity terms)}

In SPECFEM, when considering purely acoustic waves (neglecting gravity acceleration and viscosity)
in fluid regions one uses a potential formulation to represent the "density times displacement" perturbation
\cite{ChKoViCaVaFe07}:
\begin{equation}
\rho \mathbf{u} = \nabla \chi
\label{potential}
\end{equation}
where $\mathbf{u}$ is the displacement perturbation, $\rho$ the background density and $\chi$ is the scalar potential.\\
This formulation has two great advantages: it reduces the number of unknowns from three to one, and it also automatically removes artefacts
in finite-element methods (FEM) caused by the displacement formulation \cite{HaOuVe78}.
This is crucial since it is the main reason for which a displacement formulation cannot be used in fluid regions in a FEM.
\newline

However, this formulation assumes that the density is spatially constant, or locally slowly varying (e.g. inside a given spectral element),
because the use of a potential for acoustic waves is justified by the fact that acoustic displacements are irrotational.
In the cases for which the background density has strong local variations, that formulation is not valid any more.
Indeed, taking the rotational of \eqref{potential}
\begin{eqnarray*}
\nabla \times (\rho \mathbf{u}) &=& \nabla\times\nabla \chi
\end{eqnarray*}
Since for any scalar $\chi$, $ \nabla\times\nabla \chi = 0$, this yields
\begin{eqnarray*}
\nabla\rho\times\mathbf{u} + \rho\nabla\times\mathbf{u} &=& \mathbf{0}\\
\nabla\times\mathbf{u}&=&-(1/\rho)\nabla\rho\times\mathbf{u}
\end{eqnarray*}
Then, one notices that in the case of strong density gradients the rotational of $\mathbf{u}$ is no longer zero.
Thus, the potential formulation \eqref{potential} is no more valid. It is valid only when $(1/\rho)\nabla\rho\times\mathbf{u}$ can be considered negligible,
i.e. when both the gradient of density and the value of displacement are small, or if the displacement is parallel or quasi-parallel to the gradient of density,
in which case their rotational is also close to zero.
\newline

Another approach exists for acoustic waves, based on a displacement potential instead of on a potential of $\rho \mathbf{u}$:
\begin{equation}
\mathbf{u} = \nabla \chi \, ,
\end{equation}
which gives $\nabla\times\mathbf{u} = \mathbf{0}$.
But in this case the gradient of density is neglected in the Eulerian momentum equation in order to identify gradient terms.
Indeed, from the Eulerian momentum equation for inviscid acoustic waves only
\begin{eqnarray*}
\rho\partial_t^2\mathbf{u} &=& -\nabla p\\
\nabla (\partial_t^2\chi)&=& -(1/\rho)\nabla p \, .
\end{eqnarray*}
When $\nabla\rho \gg \mathbf{0}$ strictly speaking one cannot identify $\partial_t^2\chi$ with $p$.

The solution to this is to work with pressure as the scalar unknown (\cite{God11}, equation 23),
or even better with the time integral (the primitive) of pressure (see e.g. \cite{Eve81});
this was the approach that was originally implemented in SPECFEM until 10 years ago or so, but \cite{ChCaVi03} showed that switching to a displacement potential
or to a potential of $\rho \mathbf{u}$ enables one to couple the fluid and solid regions without iterating on the coupling condition in the time scheme,
thus resulting in much easier and cheaper implementation; since at that time SPECFEM users were not studying media with strong local variations of density (in solid Earth
geophysics such large local variations in fluid regions never happen), the SPECFEM developers permanently switched to such a displacement-based potential formulation.
However, for media with strong local variations of density in a fluid region we should reconsider the issue and consider trying the pressure-based formulation again.

Cases in which this can be problematic may include e.g. upper atmosphere studies: the density there is very small but displacements can become very large;
in 1D atmosphere models $\frac{\nabla \rho}{\rho}$ is approximately constant and equal to -1/H, where H is the scaling height,
which leads to $\frac{\nabla \rho}{\rho} \simeq$ 3.42*1e-5, i.e. very small; however displacements there can be large
and thus it is not clear if $(1/\rho)\nabla\rho\times\mathbf{u}$ can be considered negligible there.
